Palmer amaranth (*Amaranthus palmeri*) adaptation to US Midwest agroecosystems

### 2021-09-16 Manuscript/Manuscript.Rmd
as

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is one of the most troublesome weed species in the United States. Palmer amaranth is endemic to Southern United States but its range is expanding northward. Palmer amaranth dispersal warrants studies assessing species adaptation into new geographies. A study was conducted to investigate morphology, flowering and sex dimorphism from cohorts of Palmer amaranth growing under corn, soybean, and bareground across five locations of US Midwest. Results demonstrated that first cohort of Palmer amaranth, established in June, produced 42\% more biomass than plants from second cohort (established in July). The first Palmer amaranth cohort produced 75.5 g plant\textsuperscript{-1} in bareground, 28.3 g plant\textsuperscript{-1} in soybean and 16.3 g plant\textsuperscript{-1} in corn, whereas the second Palmer amaranth cohort produced 62.6, 6.3, and 1.4 g plant\textsuperscript{-1} in bareground, soybean and corn, respectively. Palmer amaranth height was more impacted when growing in corn. Palmer amaranth plants averaged 85.2 cm tall in the first cohort but 38.2 cm tall in the second cohort in corn. Moreover, Palmer amaranth flowering window shifted according to crop and cohort timings. Palmer amaranth growing in intense competition, such as under low light in corn, resulted in the longest flowering window. Also, Palmer amaranth sex dimorphism was slightly influenced by day of year, weight and height. The model estimated that probability of being a female plant increased as biomass and height increased. Our results showed the fast adaptation and plasticity of Palmer amaranth to grow and adapt to cropping systems from the US Midwest. Palmer amaranth is likely to continue its expansion northward. Therefore, preventing plant dispersal into new habitats is the most effective management strategy. Reactive management to reduce Palmer amaranth impact on cropping systems should encompass diversity of tactics that minimize the species ability to establish into cropping systems, including crop rotation (beyond corn and soybean), early/late crop planting, row spacing, cover crops, and effective chemical control programs.

@@

# Introduction {-}
Palmer amaranth (*Amaranthus palmeri* S. Watson) is currently ranked as one of the most economically detrimental weed species to cropping systems in the United States [@vanwychen2020]. Unmanaged Palmer amaranth plants compete for water, light, and nutrients, which can drastically impact crop yields [@berger2015]. For example, Palmer amaranth has been documented to reduce up to 91%, 68%, and 54% corn [@massinga2001], soybean [@klingaman1994], and cotton [@morgan2001] yields, respectively. Moreover, Palmer amaranth has shown a remarkable capacity to evolve resistance to herbicides. To date, Palmer amaranth has evolved resistance to eight herbicide sites of action [@heap2021], increasing the weed management complexity [@lindsay2017]. Thus, Palmer amaranth poses an economical and environmental risk to sustainable agriculture.

Palmer amaranth is a fast growing summer annual forb indigenous to the Sonoran Desert [@sauer1957]. The species would eventually emerge as a threat to US agriculture in the 1990s. Palmer amaranth weediness is likely a result of human-assisted selection in combination with plant biology. Farm mechanization, adoption of conservation agriculture (e.g., no-till), and intensive use of herbicides for weed management are the main human-mediated selection of Palmer amaranth into cropping systems [@ward2013]. On the other hand, Palmer amaranth is a prolific seed producer with a C4 photosynthetic apparatus [@wang1992]. With a dioecy nature, Palmer amaranth male and female plants are obligate outcrosser species, increasing the chances of exchanging adaptive traits among plants [@jhala2021; @oliveira2018]. Also, Palmer amaranth small seeds (e.g, 1 mm) tend to thrive in no-tillage systems [@price2011], and spread across locations through farm equipment [@sauer1972], manure [@hartzler2016], wildlife [@farmer2017], etc. The dispersal capacity of Palmer amaranth makes the species one of the most successful cases of weed adaptation to current cropping systems.

Light and temperature are the main environmental requirements for Palmer amaranth growth and development [@jha2010]. Palmer amaranth is reported with an extended germination period [@ward2013]. Germination of Palmer amaranth was triggered by 18 C soil temperature at 5 cm depth [@keeley1987], and optimal germination and biomass production occurred at 35/30 C day and night temperatures [@guo2003]. In addition, Palmer amaranth establishment is human-mediated by tillage timings and preemergence-applied herbicides [@chahal2021], which can result in weed germination shifts [@sbatella2010]. Soil water content has not shown to limit Palmer amaranth fitness. Under continuous water stress, Palmer amaranth survived and produced at least 14000 seeds plant\textsuperscript{-1} [@chahal2018]. Seeds from Palmer amaranth growing with limited water conditions were heavier, less dormant, and prompt for germination [@matzrafi2021]. Growing conditions and management practices also influence Palmer amaranth sex dimorphism and flowering pattern [@korres2017; @rumpa2019]. Therefore, Palmer amaranth has shown plasticity to evolve and fast adapt under the current agroecosystem conditions. Further scenarios show that global temperature warming can impact agriculture, and promote niches for Palmer amaranth invasion/adaptation into new environments. Currently, it is estimated that the greatest climatic risk of Palmer amaranth establishment are agronomic crops in Australia and Sub-Sahara Africa [@kistner2018]. Temperature is a key factor limiting Palmer amaranth expansion to cooler geographies [@briscoerunquist2019]; however, under future climate change Palmer amaranth is likely to expand northward into Canada and Northern Europe [@kistner2018; @briscoerunquist2019].

Palmer amaranth is already found in agronomic crops of South America [@larran2017; @kupper2017] and Southern Europe [@milani2021]. In the United States, Palmer amaranth is well established in the Cotton Belt [@garetson2019; @bagavathiannan2016] in the southern United States but its range is expanding northward. For example, herbicide resistant Palmer amaranth is widespread in Nebraska  [@oliveira2021a]. There are some reported cases of Palmer amaranth in Michigan [@kohrt2017] and Connecticut [@aulakh2021]. Also, it is estimated that Palmer amaranth can cause high damage to soybean fields in Illinois [@davis2015], which is concerning as soybean along with corn make most of US Midwest agronomic hectares. In Iowa, a study showed that Palmer amaranth is still not well adapted compared to waterhemp (*Amaranthus tuberculatus*) [@baker2021]. Waterhemp, an *Amaranthus* species, is the most troublesome species in the US Midwest [@tranel2011]. Nonetheless, invasion and successful eradication of Palmer amaranth is documented in Minnesota [@yu2021]. Palmer amaranth infestations have not been detected in Canada; however, Palmer amaranth seeds were detected in sweet potato slips in the country [@page2021]. Palmer amaranth is still not as well adapted and established to Northern as it is in the Southern North America. Therefore, Palmer amaranth range of expansion into new habitats can increase. It seems certain the need to manage new Palmer amaranth infestations in agronomic crops throughout northern United States in the near future. Strategies on Palmer amaranth management should encompass the agroecosystem level but not only attempts to eradicate the weed. Most tactics to manage Palmer amaranth are based on technology fixes [@scott2011], which are short-term (e.g., herbicide and/or tillage) rather than long-term weed management. Palmer amaranth management should be built on minimizing the species ability to adapt, grow and develop into agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to have a higher impact on cotton yields compared to late established plants [@macrae2013]. In the northern states, Palmer amaranth impact on the agroecosystem is recent. Studies investigating Palmer amaranth in those locations are limited due to the plant classification as noxious weed species in some northern states [@yu2021]. Nonetheless, the continuous Palmer amaranth dispersal and potential establishment across the northern United States is concerning and warrants investigations on species morphology in such environments. Understanding Palmer amaranth biology and growing strategies under different agroecosystems can enhance our knowledge on species adaptation and management practices. It can also aid in designing proactive and ecological tactics to limit the species range expansion, reduce its negative impact, and developing resilient and sustainable farming systems [@maclaren2020]. Therefore, the objective of this study was to investigate the flowering pattern, sex dimorphism, biomass production, and height of Palmer amaranth cohorts growing under corn, soybean and bareground across five locations in the United States Midwest.

@@

An ecological approach to reduce Palmer amaranth seed production is understanding plant biology, including flowering pattern. Our study suggests that Palmer amaranth flowering was slightly influenced by crops and cohort timings. Palmer amaranth growing in bareground and corn resulted in the overall shortest and longest flowering window, respectively. The shortest flowering window of second cohorts of Palmer amaranth growing in bareground highlighted the impact of photoperiod on flowering. When growing in soybean, Palmer amaranth flowering window was similar to bareground at first cohort but similar to corn at second cohort timing. Plant flowering initiation is complex and depends on biological and ecological factors [@lang1965]. We hypothesize that when growing in high competition (e.g., second cohort), Palmer amaranth plants tend to initiate flowering early, as well as having an extended flowering window. A study has shown that Palmer amaranth initiated flowering two weeks prior to the native waterhemp in Iowa [@baker2021]. Early flower initiation could also be a plant strategy when growing in stress conditions. For example, when growing under water stress, early flowering in Palmer amaranth resulted in a mismatch between female and male plants by seven days [@mesgaran2021]. A mismatch in Palmer amaranth male and female flowering period can minimize plant outcross, reducing plant seed production and/or exchange of resistant alleles [@jhala2021]. Sex dimorphism manipulation is considered a potential ecological pest control [@mcfarlane2018; @schliekelman2005].

@@

Our study demonstrated the Palmer amaranth plasticity to grow and develop into arable land of US Midwest. It is likely that Palmer amaranth range will continue to expanding into new geographies. The migration of Palmer amaranth into the US Midwest will reshape the landscape as waterhemp and Palmer amaranth will share the same habitat. The presence of Palmer amaranth and waterhemp will increase the weed management complexity. Therefore, preventive management is a priority to minimizing Palmer amaranth dispersal. Reactive management should focus on early-season management programs, which would have a large negative effect on Palmer amaranth growth and development. Long-term tactics that promote early-season crop advantage against Palmer amaranth, including diversity of crops in rotation, early/late crop planting, plant width, and crop residue (e.g. cover crops) would minimize the negative impact of Palmer amaranth to cropping systems. The aggressiveness and differential Palmer amaranth adaptation to agroecosystem is striking and require national efforts to minimize the species impact on sustainability and profitability of cropping systems.

@@

RW and MO: designed the experiments; AJ, CP, MB, MO, and SS: conducted the experiments; MO: analyzed the data and wrote the manuscript; AJ, CP, MB, MO, SS, and RW: conceptualized the research. All authors reviewed the manuscript.

>>>>

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is one of the most troublesome agronomic weed species in the United States. Palmer amaranth is endemic to Southern United States but its range is expanding northward. Palmer amaranth dispersal warrants studies assessing species adaptation into new geographies. A study was conducted to investigate morphology, flowering and gender from cohorts of Palmer amaranth growing under corn, soybean, and bareground across five locations of US Midwest. Results demonstrated that first cohort of Palmer amaranth, established in June, produced 42\% more biomass than plants from second cohort (established in July). The first Palmer amaranth cohort produced 75.5 g plant\textsuperscript{-1} in bareground, 28.3 g plant\textsuperscript{-1} in soybean and 16.3 g plant\textsuperscript{-1} in corn, whereas the second Palmer amaranth cohort produced 62.6, 6.3, and 1.4 g plant\textsuperscript{-1} in bareground, soybean and corn, respectively. Palmer amaranth height was more impacted when growing in corn. Palmer amaranth plants averaged 85.2 cm tall in the first cohort but 38.2 cm tall in the second cohort in corn. Moreover, Palmer amaranth flowering window shifted according to crop and cohort timings. Palmer amaranth growing in intense competition, such as under low light in corn, resulted in the longest flowering window. Also, Palmer amaranth sex dimorphism was slightly influenced by day of year, weight and height. The model estimated that probability of being a female plant increased as biomass and height increased. Our results showed the fast adaptation and plasticity of Palmer amaranth to grow and adapt to cropping systems from the US Midwest. Palmer amaranth is likely to continue its expansion northward. Therefore, preventing plant dispersal into new habitats is the most effective management strategy. Reactive management to reduce Palmer amaranth impact on cropping systems should encompass diversity of tactics that minimize the species ability to establish into cropping systems, including crop rotation (beyond corn and soybean), early/late crop planting, row spacing, cover crops, and effective chemical control programs.

@@

# Introduction {-}
Palmer amaranth (*Amaranthus palmeri* S. Watson) is currently ranked as one of the most economically detrimental weed species to cropping systems in the United States [@vanwychen2020]. Unmanaged Palmer amaranth plants compete for water, light, and nutrients, which can drastically impact crop yields [@berger2015]. For example, Palmer amaranth has been documented to reduce up to 91%, 68%, and 54% corn [@massinga2001], soybean [@klingaman1994], and cotton [@morgan2001] yields, respectively. Moreover, Palmer amaranth has shown a remarkable capacity to evolve resistance to herbicides. To date, Palmer amaranth has evolved resistance to eight herbicide sites of action [@heap2021], increasing the weed management complexity [@lindsay2017]. Thus, Palmer amaranth poses an economical and ecological risk to sustainable agriculture.

Palmer amaranth is a fast growing summer annual forb indigenous to the Sonoran Desert [@sauer1957]. The species would eventually emerge as a threat to US agriculture in the 1990s. Palmer amaranth weediness is likely a result of human-assisted selection in combination with plant biology. Farm mechanization, adoption of conservation agriculture (e.g., no-till), and intensive use of herbicides for weed management are the main human-mediated selections of Palmer amaranth into cropping systems [@ward2013]. On the other hand, Palmer amaranth is a prolific seed producer with a C4 photosynthetic apparatus [@wang1992]. With a dioecious nature, Palmer amaranth male and female plants are obligate outcrosser species, increasing the chances of exchanging adaptive traits among plants [@jhala2021; @oliveira2018]. Also, Palmer amaranth's small seeds (e.g, 1 mm) tend to thrive in no-tillage systems [@price2011], and spread across locations through farm equipment [@sauer1972], seed mixes [@hartzler2016], wildlife [@farmer2017], etc. The dispersal capacity of Palmer amaranth makes the species one of the most successful cases of weed adaptation to current cropping systems.

Light and temperature are the main environmental requirements for Palmer amaranth growth and development [@jha2010]. Palmer amaranth is reported with an extended germination period [@ward2013]. Germination of Palmer amaranth was triggered by 18 C soil temperature at 5 cm depth [@keeley1987], and optimal germination and biomass production occurred at 35/30 C day and night temperatures [@guo2003]. In addition, Palmer amaranth emergence is human-mediated by tillage timings and preemergence-applied herbicides [@chahal2021], which can result in weed germination shifts [@sbatella2010]. Also, Palmer amaranth has shown to grow and develop under reduced soil water content. In continuous water stress, Palmer amaranth survived and produced at least 14000 seeds plant\textsuperscript{-1} [@chahal2018]. Seeds from Palmer amaranth growing with limited water conditions were heavier, less dormant, and prompt for germination [@matzrafi2021]. Growing conditions and management practices also influence Palmer amaranth sex dimorphism and flowering pattern [@korres2017; @rumpa2019]. Therefore, Palmer amaranth has shown plasticity to evolve and fast adapt under the current agroecosystem conditions. Future scenarios suggest that global temperature warming can impact agriculture, and promote niches for Palmer amaranth invasion/adaptation into new environments. Agronomic crops in Australia and Sub-Sahara Africa are currently at greatest risk to Palmer invasion as climate warms [@kistner2018]. Temperature is a key factor limiting Palmer amaranth expansion to cooler geographies [@briscoerunquist2019]; however, under future climate change Palmer amaranth is likely to expand northward into Canada and Northern Europe [@kistner2018; @briscoerunquist2019].

Palmer amaranth is already found in agronomic crops of South America [@larran2017; @kupper2017] and Southern Europe [@milani2021]. In the United States, Palmer amaranth is well established in the Cotton Belt [@garetson2019; @bagavathiannan2016] in the southern United States but its range is expanding northward. For example, herbicide resistant Palmer amaranth is widespread in Nebraska  [@oliveira2021a]. There are some reported cases of Palmer amaranth in Michigan [@kohrt2017] and Connecticut [@aulakh2021]. Also, it is estimated that Palmer amaranth can cause yield loss to soybean fields in Illinois [@davis2015], which is concerning as soybean along with corn comprise most of US Midwest agronomic hectares. In Iowa, a study showed that Palmer amaranth is still not well adapted compared to waterhemp (*Amaranthus tuberculatus*) [@baker2021], which is considered the most troublesome species in the US Midwest [@tranel2011]. Invasion and successful eradication of Palmer amaranth is documented in Minnesota [@yu2021]. Palmer amaranth infestations have not been detected in Canada; however, Palmer amaranth seeds were detected in sweet potato slips in the country [@page2021]. Palmer amaranth is still not as well adapted and established to Northern as compared to Southern North America. Therefore, Palmer amaranth range of expansion into new habitats can increase. It seems certain the need to manage new Palmer amaranth infestations in agronomic crops throughout northern United States is eminent in the near future. Strategies on Palmer amaranth management should encompass the agroecosystem level but not focus only on attempts to eradicate the weed. Most tactics to manage Palmer amaranth are based on technological solutions [@scott2011], which are short-term (e.g., herbicide and/or tillage) rather than long-term weed management. Palmer amaranth management should be built around minimizing the species ability to adapt, grow and develop successfully in agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to have a higher impact on cotton yields compared to late established plants [@macrae2013]. In the northern states, Palmer amaranth impact on the agroecosystem is recent. Studies investigating Palmer amaranth in northern states are limited due to the plant classification as noxious weed species [@yu2021]. Nonetheless, the continuous Palmer amaranth dispersal and potential establishment across the northern United States is concerning and warrants investigations on species morphology in northern environments. Understanding Palmer amaranth biology and morphology under different agroecosystems can enhance our knowledge on species adaptation and management practices. It can also aid in designing proactive and ecological tactics to limit the species range expansion, reduce its negative impact on agroecossystems, and to develop resilient and sustainable farming systems [@maclaren2020]. Therefore, the objective of this study was to investigate the flowering pattern, sex dimorphism, biomass production, and height of Palmer amaranth cohorts growing under corn, soybean and bareground across five locations in the United States Midwest.

@@

An ecological approach to reduce Palmer amaranth seed production is understanding plant biology, including flowering pattern. Our study suggests that Palmer amaranth flowering was slightly influenced by crops and cohort timings. Palmer amaranth growing in bareground and corn resulted in the overall shortest and longest flowering window, respectively. The shortest flowering window of second cohorts of Palmer amaranth growing in bareground highlighted the impact of photoperiod on flowering. When growing in soybean, Palmer amaranth flowering window was similar to bareground at first cohort but similar to corn at second cohort timing. Plant flowering initiation is complex and depends on the complex interaction between genetic makeup and the environmental conditions [@lang1965]. We hypothesize that when growing in high competition (e.g., second cohort), Palmer amaranth plants tend to initiate flowering early, as well as having an extended flowering window. A study has shown that Palmer amaranth initiated flowering two weeks prior to the native waterhemp in Iowa [@baker2021]. Early flower initiation could also be a plant strategy when growing in stress conditions. For example, when growing under water stress, early flowering in Palmer amaranth resulted in a mismatch between female and male plants by seven days [@mesgaran2021]. A mismatch in Palmer amaranth male and female flowering period can minimize plant outcross, reducing plant seed production and/or exchange of resistant alleles [@jhala2021]. Sex dimorphism manipulation is considered a potential ecological pest control [@mcfarlane2018; @schliekelman2005].

@@

Our study demonstrated the Palmer amaranth plasticity to grow and develop into arable land of US Midwest. It is likely that Palmer amaranth range will continue to expanding into new geographies. The migration of Palmer amaranth into the US Midwest will reshape the landscape as waterhemp and Palmer amaranth will share the same habitat. The presence of Palmer amaranth and waterhemp will increase the weed management complexity. Therefore, preventive management is a priority to minimizing Palmer amaranth dispersal. Reactive management should focus on early-season management programs, which would have a large negative effect on Palmer amaranth growth and development. Long-term tactics that promote early-season crop advantage against Palmer amaranth, including diversity of crops in rotation, early/late crop planting, plant width, and crop residue (e.g. cover crops) would minimize the negative impact of Palmer amaranth to cropping systems. The aggressiveness and differential Palmer amaranth adaptation to agroecosystem is striking and require regional collaboration to minimize the species impact on sustainability and profitability of cropping systems.

@@

RW and MO: designed the experiments; AJ, CP, MB, MO, and SS: conducted the experiments; MO: analyzed the data; MO: wrote the manuscript and all authors revised; AJ, CP, MB, MO, SS, and RW: conceptualized the research. All authors reviewed the manuscript.
